A sheaf--function dictionary for algebraic tori over local fields

### 2013-10-02 shanghai.tex
Edits, mainly to section 5

Consider the short exact sequence in the category of $\QC$-schemes
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \rar & G \rar & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is also a $\QC$-scheme, 
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines a sequence of functors
\[
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \rar & \QC(G) \rar & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\]
which induces a sequence of group homomorphisms 
\[
\begin{tikzcd}
\QCiso{0} \rar & \QCiso{\pi_0(G)} \rar & \QCiso{G} \rar & \QCiso{G^0} \rar & \QCiso{0}.
\end{tikzcd}
\]
after passing to isomorphism classes.  It is easy to see that $\QCiso{0}$ is trivial. In Sections~\ref{sec:connected}, \ref{sec:etale} and \ref{sec:etalecohom} we will find $\QCiso{G^0}$
and $\QCiso{\pi_0(G)}$ under the hypothesis that $\pi_0(G)$ is geometrically finitely generated; 
the snake lemma will then allow us to show that
\[
\QCiso{G} \iso G(\Fq)^*.
\]

@@

Now, let $\qcs{L} \ceq (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$. 
Set $\qcs{L}_1 \ceq q_1^* \qcs{L}$ and $\qcs{L}_2 \ceq q_2^* \qcs{L}$;
by Proposition~\ref{prop:pullback}, $\qcs{L}_1$ a quasicharacter sheaf on $G_1$ 
and $\qcs{L}_2$ is a quasicharacter sheaf on $G_2$.
We will obtain an isomorphism $\qcs{L} \iso  \qcs{L}_1\boxtimes \qcs{L}_2$.
%
Applying the functor $f^*$ to the isomorphism $\mu$ yields
\begin{equation}\label{eq:fm}
f^*\mu : f^* m^* \gqcs{L} \to f^*(\gqcs{L}\boxtimes \gqcs{L}) .
\end{equation}
We have already observed that $m\circ f = \id_G$, so $f^* m^* \gqcs{L} = \gqcs{L}$.
On the other hand, $f^*(\gqcs{L}\boxtimes \gqcs{L}) = f^*p_1^*\gqcs{L}\otimes f^* p_2^*\gqcs{L}$, 
by definition, and $f^*p_1^*\gqcs{L} = r_1^* q_1^* \gqcs{L}$ and $f^*p_2^*\gqcs{L} = r_2^* q_2^* \gqcs{L}$ follows from the properties of these maps, articulated above, 
so $f^*(\gqcs{L}\boxtimes \gqcs{L}) = \gqcs{L}_1\boxtimes \gqcs{L}_2$.
It follows that \eqref{eq:fm} gives an isomorphism $\gqcs{L} \to  \gqcs{L}_1\boxtimes \gqcs{L}_2$.
It is a matter of routine to show that this morphism satisfies Conditions~\ref{CS.2} as they apply here,
from which it follows that we have exhibited an isomorphism 
$\qcs{L} \to \qcs{L}_1\boxtimes \qcs{L}_2$ of quasicharacters sheaves on $G\times G$.
This completes the proof that $\boxtimes$ is essentially surjective.
\end{proof}

@@

If $\qcs{L} \ceq (\gqcs{L}, \mu, \phi)$ is a quasicharacter sheaf on $G$ we may define
a quasicharacter sheaf $\qcs{L}' \ceq (\gqcs{L}, \mu, \phi_{k'})$ on the base change
$G_{k'}$ of $G$ to $k'$ by setting
\[
\phi_{k'} \ceq \phi \circ \Frob{G}^*(\phi) \circ \cdots \circ (\Frob{G}^{n-1})^*(\phi).
\]
The commutativity of the diagram (CS.2) for $\phi_{k'}$ 
follows from the fact that $\Frob{G_{k'}} = \Frob{G}^n$.
Note that we may also think about the construction of $\phi_{k'}$ from $\phi$
as taking the action $\varphi$ of $\Weil{k}$ on $\gqcs{L}$ 
defined in Section~\ref{sec:category} and restricting it to the subgroup $\Weil{k'}$.

@@

\begin{definition}
Let $\QCb(G)$ be the category of pairs $(\qcs{L}_0,\mu_0)$ 
where $\qcs{L}_0$ an $\ell$-adic local system on $G$ of rank~one, 
equipped with an isomorphism $\mu_0 : m^* \qcs{L}_0 \to \qcs{L}_0 \boxtimes \qcs{L}_0$ 
satisfying the analogue of Condition~\ref{CS.1} on $G$; 
morphisms in $\QCb(G)$ are defined as in (the second part of) Condition~\ref{CS.2}. 
This is the category of \emph{bounded quasicharacter sheaves} on $G$. 
\end{definition}

@@

\begin{proof}
 Let $b_G : {\bar G} \to G$ be the pull-back of $\Spec{\bFq} \to \Spec{\Fq}$ along $G\to \Spec{\Fq}$.
 Let $(\qcs{L}_0,\mu_0)$ be a bounded quasicharacter sheaf on $G$. 
 Then $\qcs{L}_0$ is an $\ell$-adic constructible sheaf on $G$ and
 $b_G^* \qcs{L}_0$ comes equipped with an isomorphism 
 $\phi : \Frob{G}^* b_G^*\qcs{L}_0 \to b_G^* \qcs{L}_0$.
 Moreover, the functor $\qcs{L}_0 \mapsto (b_G^* \qcs{L}_0,\phi)$  
 from $\ell$-adic constructible sheaves on $G$ to $\ell$-adic constructible sheaves on $G$
 is full and faithful \citelist{\cite{deligne-katz:SGA7.2}*{Expos\'e XIII} \cite{beilinson-bernstein-deligne:81a}*{Prop. 5.2.1}}. 
 This functor preserves local constancy, so takes local systems to local systems. 
 Set $\mu \ceq b_{G\times G}^*\mu_0$; clearly, this satisfies Condition~\ref{CS.2} 
 with $b_G^*\qcs{L}_0$ playing the role of $\gqcs{L}$.
 Moreover, $\phi$ is compatible with $\mu$ in the sense of Condition~\ref{CS.3}.
 This defines the functor $B_G^* : (\qcs{L}_0,\mu_0) \mapsto (b_G^*\qcs{L}_0,\mu, \phi)$
 and also shows that it is full and faithful.
\end{proof}

In general, $\QCb(G)$ is an essentially
proper subcategory of $\QC(G)$. Indeed, it follows from Theorem~\ref{thm:snake}
that the restriction of $\trFrob{G} : \QCiso{G} \to G(\Fq)^*$ to the subgroup
$\QCbiso{G}$ is an isomorphism onto the subgroup of $\chi \in G(\Fq)^*$ such
that the image of $\chi$ is bounded in $\EEx$; 
see Section~\ref{sec:snake} for more details regarding this point.

>>>>

Consider the short exact sequence in the category of $\QC$-schemes
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \rar & G \rar & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is also a $\QC$-scheme, 
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines a sequence of functors
\[
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \rar & \QC(G) \rar & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\]
which induces a sequence of group homomorphisms 
\[
\begin{tikzcd}
\QCiso{0} \rar & \QCiso{\pi_0(G)} \rar & \QCiso{G} \rar & \QCiso{G^0} \rar & \QCiso{0}.
\end{tikzcd}
\]
after passing to isomorphism classes.  It is easy to see that $\QCiso{0}$ is trivial. In Sections~\ref{sec:connected} and \ref{sec:etale} we will find $\QCiso{G^0}$
and $\QCiso{\pi_0(G)}$ under the hypothesis that $\pi_0(G)$ is geometrically finitely generated; 
the snake lemma will then allow us to show that
\[
\QCiso{G} \iso G(\Fq)^*.
\]

@@

Now, let $\qcs{L} \ceq (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$. 
Set $\qcs{L}_1 \ceq q_1^* \qcs{L}$ and $\qcs{L}_2 \ceq q_2^* \qcs{L}$;
by Proposition~\ref{prop:pullback}, $\qcs{L}_1$ a quasicharacter sheaf on $G_1$ 
and $\qcs{L}_2$ is a quasicharacter sheaf on $G_2$.
We will obtain an isomorphism $\qcs{L} \iso  \qcs{L}_1\boxtimes \qcs{L}_2$.
%
Applying the functor $f^*$ to the isomorphism $\mu$ yields
\begin{equation}\label{eq:fm}
f^*\mu : f^* m^* \gqcs{L} \to f^*(\gqcs{L}\boxtimes \gqcs{L}) .
\end{equation}
We have already observed that $m\circ f = \id_G$, so $f^* m^* \gqcs{L} = \gqcs{L}$.
On the other hand, $f^*(\gqcs{L}\boxtimes \gqcs{L}) = f^*p_1^*\gqcs{L}\otimes f^* p_2^*\gqcs{L}$
by definition.  Since $f^*p_1^*\gqcs{L} = r_1^* q_1^* \gqcs{L}$ and $f^*p_2^*\gqcs{L} = r_2^* q_2^* \gqcs{L}$
we get that $f^*(\gqcs{L}\boxtimes \gqcs{L}) = \gqcs{L}_1\boxtimes \gqcs{L}_2$.
It follows that \eqref{eq:fm} gives an isomorphism $\gqcs{L} \to  \gqcs{L}_1\boxtimes \gqcs{L}_2$.
It is a matter of routine to show that this morphism satisfies Conditions~\ref{CS.2} as they apply here,
from which it follows that we have exhibited an isomorphism 
$\qcs{L} \to \qcs{L}_1\boxtimes \qcs{L}_2$ of quasicharacters sheaves on $G\times G$.
This completes the proof that $\boxtimes$ is essentially surjective.
\end{proof}

@@

If $\qcs{L} \ceq (\gqcs{L}, \mu, \phi)$ is a quasicharacter sheaf on $G$, we may define
a quasicharacter sheaf $\qcs{L}' \ceq (\gqcs{L}, \mu, \phi_{k'})$ on the base change
$G_{k'}$ of $G$ to $k'$ by setting
\[
\phi_{k'} \ceq \phi \circ \Frob{G}^*(\phi) \circ \cdots \circ (\Frob{G}^{n-1})^*(\phi).
\]
The commutativity of the diagram (CS.2) for $\phi_{k'}$ 
follows from the fact that $\Frob{G_{k'}} = \Frob{G}^n$.
Note that we may also think about the construction of $\phi_{k'}$ from $\phi$
as taking the action $\varphi$ of $\Weil{k}$ on $\gqcs{L}$ 
defined in Section~\ref{sec:category} and restricting it to the subgroup $\Weil{k'}$.

@@

\begin{definition}
Let $\QCb(G)$ be the category of pairs $(\qcs{L}_0,\mu_0)$ 
where $\qcs{L}_0$ an $\ell$-adic local system on $G$ of rank~one, 
equipped with an isomorphism $\mu_0 : m^* \qcs{L}_0 \to \qcs{L}_0 \boxtimes \qcs{L}_0$ 
satisfying the analogue of Condition~\ref{CS.1} on $G$; 
morphisms in $\QCb(G)$ are defined as in the second part of Condition~\ref{CS.2}. 
This is the category of \emph{bounded quasicharacter sheaves} on $G$. 
\end{definition}

@@

\begin{proof}
 Let $b_G : {\bar G} \to G$ be the pull-back of $\Spec{\bFq} \to \Spec{\Fq}$ along $G\to \Spec{\Fq}$.
 Let $(\qcs{L}_0,\mu_0)$ be a bounded quasicharacter sheaf on $G$. 
 Then $\qcs{L}_0$ is an $\ell$-adic constructible sheaf on $G$ and
 $b_G^* \qcs{L}_0$ comes equipped with an isomorphism 
 $\phi : \Frob{G}^* b_G^*\qcs{L}_0 \to b_G^* \qcs{L}_0$.
 Moreover, the functor $\qcs{L}_0 \mapsto (b_G^* \qcs{L}_0,\phi)$  
 from $\ell$-adic constructible sheaves on $G$ to $\ell$-adic constructible sheaves on $G$
 is full and faithful \citelist{\cite{deligne-katz:SGA7.2}*{Expos\'e XIII} \cite{beilinson-bernstein-deligne:81a}*{Prop. 5.2.1}}. 
 This functor preserves local constancy, so takes local systems to local systems. 
 Set $\mu \ceq b_{G\times G}^*\mu_0$; clearly, this satisfies Condition~\ref{CS.2} 
 with $b_G^*\qcs{L}_0$ playing the role of $\gqcs{L}$.
 Moreover, $\phi$ is compatible with $\mu$ in the sense of Condition~\ref{CS.3}.
 This construction defines the functor $B_G^* : (\qcs{L}_0,\mu_0) \mapsto (b_G^*\qcs{L}_0,\mu, \phi)$
 and also shows that it is full and faithful.
\end{proof}

In general, $\QCb(G)$ is an essentially
proper subcategory of $\QC(G)$. Indeed, it follows from Theorem~\ref{thm:snake}
that the restriction of $\trFrob{G} : \QCiso{G} \to G(\Fq)^*$ to the subgroup
$\QCbiso{G}$ is an isomorphism onto the subgroup of $\chi \in G(\Fq)^*$ such
that the image of $\chi$ is bounded in $\EEx$; 
see Section~\ref{sec:snake} for more details.
